Tropical Mathematics and the Lambda Calculus

### 2023-01-22 section3new.tex
commit

The study of matrices with values over the tropical semi-ring can be seen as a special case of the
\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus.

@@

The tropical relational semantics is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
It is worth observing that the formula for composition in $\LREL$ is the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
 similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
%\end{remark}

@@

>>>>

\subsection{Tropical linear algebra% and computation
}

At the basis of our approach is the observation that the \emph{tropical semiring} $([0,\infty], \min, +)$, which is at the heart of tropical mathematics, coincides with the \emph{Lawvere quantale} $\Lawv=([0,\infty], \geq, +)$ \cite{Hofmann2014, Stubbe2014}, the structure at the heart of the categorical study of metric spaces initiated by Lawvere himself \cite{Lawvere1973}.
Let us recall that a quantale is a complete lattice endowed with a continuous monoid action. In the case of $\Lawv$ the lattice is defined by the reverse order $\geq$ on $\BB R$, and the monoid action is provided by addition. Notice that the lattice join operation of $\Lawv$ coincides with the idempotent semiring operation $\min$. 
A consequence of these observations is that, as we discussed below, the tropical approach to linear algebra coincides with the study of ``$\Lawv$-valued matrices'', i.e.~of maps of the form $s: X\times Y\to \Lawv$ .
In particular, a (possibly $\infty$) metric on a set $X$ is nothing but a ``$\Lawv$-valued square matrix'' $d:X\times X\to \Lawv$ satisfying axioms like e.g.~the triangular law (indeed, such distance matrices correspond to $\Lawv$-\emph{enriched categories}, a viewpoint we explicitly take in Section \ref{section6}).

The study of matrices with values over the tropical semiring can be seen as a special case of the
\emph{quantitative relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus.

@@

The tropical relational model is thus provided by the category $\LREL$ of matrices with values over $\Lawv$ (which, being a quantale, is indeed a continuous semi-ring).
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
It is worth observing that the formula for composition in $\LREL$ is the tropicalisation of the one defining it in $\QREL$, i.e.\ $(st)_{a,c}:=\inf\limits_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
 similarly, the linear functions $f:\Lawv^X\to \Lawv^Y$ induced by matrices, which we call \emph{tropical linear}, are exactly those of shape $f(x)=\inf\limits_{b\in Y} \set{\matr f_{a,b}+x_a}$, for some matrix $\matr f$ from $X$ to $Y$.
%\end{remark}

@@

%
%
%
%

%
%
%or more generally as a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ with coefficient $\widehat f_{n}\in [0,1]$, we can define its \emph{tropicalization} $\trop f: \Lawv \to \Lawv$ as the function 
%\begin{align}
%\trop f(\alpha)= \inf_{n}\left\{ 
%\end{align}
%
%This correspondence can be made precise through the the so-called
% \emph{de Maslov dequantization} \cite{}.
% For each positive real $t$, any polynomial in $\BB R[x]$ can be written under the \emph{$t$-parameterized} form:
% \begin{align}
% p_{t}(x)= \sum_{i=1}^{k}t^{c_{i}}x^{i}
% \end{align}
% with the coefficients $c_{i}$ taken from $\Lawv$. 
% It is clear then that tropical polynomials and $t$-parameterized polynomials admit a one to one correspondence between their presentations.
% 
% Actually, the $\varphi$s and the $p_{t}$s can be related by passing through some intermediate functions $\varphi_{t}$ introduced by Maslov.
%For any $t>1$, the functions $\phi_{t}(x)=-\log_{t}x$ and $\psi_{t}(\alpha)=t^{-\alpha}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
%

%
%{\color{red}Lista delle cose da dire:}
%
%1) Def di quantale, come lattice e come complete idempotent semiring.
%Among the the so-called \emph{tropical semirings}, we consider the \emph{Lawvere quantale/semiring}.
%
%2) Def di $\Lawv$, the \emph{Lawvere quantale}: seen as the idempotent complete semiring, it is $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,\cdot,0)$.
%Seen as lattice it is defined by the order $\preceq$, which is the reversed order $\geq$ of the usual order $\leq$ on $\BB R_{\geq0}\cup\set{\infty}$.
%
%3) Maslov dequantisation:
%
%First, let us recall that for any non-negative real $t$, the functions $\phi_{t}(x)=-t\log x$ and $\psi_{t}(\alpha)=e^{-\alpha/t}$ are inverse of each other and define thus continuous (w.r.t.\ the usual topologies) bijections between the space of probabilities $[0,1]$ and $\BB R_{\geq 0}\cup\set{\infty}$ (we write $\log$ for the natural logarithm).
%Moreover, if we set $\alpha \widetilde+ \beta:= \frac{\alpha+\beta}{2}$, $\alpha\sumt{t}\beta=\phi_{t}(\psi_{t}(\alpha)\widetilde{+}\psi_{t}(\beta))=-\log (e^{-\alpha/t}+e^{-\beta/t})-\phi_{t}(2)$ and $\alpha\prodt{t} \beta:=\phi_{t}(\psi_{t}(\alpha)\psi_{t}(\beta))=\alpha+\beta$, it is known that: $\lim\limits_{t\to 0}\alpha\sumt{t} \beta= \min\{\alpha,\beta\}$.
%In this sense, setting $\Lawv_t:=([0,\infty],\sumt{t},\prodt{t})$, one says that $\Lawv_t\to_{t\to 0^+}\Lawv$.
%Moreover, setting $\widetilde\Lawv:=([0,\infty],\widetilde+,\cdot)$, it can be shown that $\Lawv_t\simeq\widetilde\Lawv$ for all $t>0$, so the $\Lawv_t$ are all isomorphic, whereas at the limit we have a discontinuity: it can be shown that $\Lawv_t\not\simeq\Lawv$.
%
%4) Def di $\trop$ di un polinomio/serie, (\`e la stessa formula, dipende solo se gli indici sono finiti/infiniti).
%Come sta scritto sotto, giusto un po' pi\`u formale (per esempio, scriverlo come Definizione).
%
%The fundamental observation that led to the study of mathematics over the \emph{tropical semi-ring} $\Lawv=([0,\infty],\min,+)$ was that, by replacing everywhere the ``$+$'' by the ``$\min$'' and the ``$\times$'' by the ``$+$'', many algebraic and geometric objects becomes combinatorial and their computation simpler. 
%
%For instance, the tropicalization of a cubic polynomial $p(x)=ax^{3}+bx^{2}+cx+d$ yields a piecewise-linear function 
%\begin{align}
%\trop p(\alpha)= \min\{ 3\alpha+a, 2\alpha+b, \alpha+c,d\}
%\end{align}
%Notably, the \emph{tropical roots} (whose definition is recalled in Section \ref{section3}) of $\trop p(\alpha)$ can be found through a rather simple (indeed polytime \cite{}) algorithm, and can be used to \emph{approximate} the actual roots of $p(x)$ \cite{}. 
%More generally, the tropicalization of a power series $f(x)=\sum_{n}\widehat f_{n}x^{n}$ yields a \emph{tropical Laurent series} \cite{} 
%\begin{align}
%\trop f(\alpha)= \inf_{n}\left\{n\alpha+ \widehat f_{n}\right\}
%\end{align}
%a class of functions that we will study in detail in Section \ref{section4}.
%
%%- generalities about tropical maths (tropicalisation $\trop P$  of polynomials and of Laurent series, and their roots -- all that without $\LREL$)
%
%
%

### 2023-01-22 section4.tex
commit

A tropical polynomial, as we saw, is a piece-wise linear function of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$. For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

As anticipated in Section \ref{sectionnew}, a tropical Laurent series (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$. In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree. For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example, 
is the ``limit'' of the polynomials $\varphi_{n}$. Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials. For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}
Call $\leq$ the (well-founded) pointwise order on $\N^k$.
Now \mbox{$\C F_\epsilon:=\set{n\in\N \mid
\widehat f(n)<\infty \textit{ and } 
\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}
does the job.
\end{proof}

@@

Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
\begin{proposition}
For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{\B x\in \Lawv^{X}}
\norm{ f(\B x)-g(\B x)}_{\infty}$.\end{proposition}

>>>>

A tropical polynomial iss a piece-wise linear function of the form \begin{equation}
f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}
\end{equation}
For example, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

A \emph{tropical Laurent series} (of one variable $x\in\Lawv$), shortly a \emph{tLs}, is a function that can be expressed as $f(x)=\inf_{n}\{nx+\matr f_{n}\}$, with $\matr  f_{n}$ a sequence in $\Lawv$. In other words, a tLs is a ``limit'' of tropical polynomials of higher and higher degree. For instance, the function $\varphi(x):=\inf_{n\in\N}\set{nx+\frac{1}{2^n}}$ (illustrated in Fig.~\ref{fig:plot1}), that we will take as our running example, 
is the ``limit'' of the polynomials $\varphi_{n}$. Since $\inf$s are not in general $\min$s, the behavior of tLS may be less predictable than that of tropical polynomials. For instance, tropical roots for tLs (see \cite{Porzio2021}) may also include limit points.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}[Proof sketch]
\mbox{$\C F_\epsilon:=\set{n\in\N \mid
\widehat f(n)<\infty \textit{ and } 
\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}, where $\leq$ is the pointwise order on $\N^k$.
The core of the proof is showing that this set is indeed finite and enough for computing $f$.
\end{proof}

@@

%Before looking at what happens in the case of non-linear programs, let us make the metric structure of $\LREL$ explicit. 
The following proposition provides a useful characterization of the functional metrics in $\LREL$, relying on 
the bijection between $\HOM{\LREL}{X}{Y}$ and set of tropical linear functions from $\Lawv^{X}$ to $\Lawv^{Y}$.
\begin{proposition}
For all tropical linear functions $f,g:\Lawv^{X}\to \Lawv^{Y}$, $\norm{ \matr f-\matr g}_{\infty} =  \sup_{\B x\in \Lawv^{X}}
\norm{ f(\B x)-g(\B x)}_{\infty}$.\end{proposition}

### 2023-01-22 section5.tex
commit

In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs. 
The general pattern is to consider a program $M$, which can be expressed in the form of a polynomial or a power series, and 
investigate what quantitative properties of $M$ can be deduced from properties of its tropicalization $\trop M$ (which are often easier to study).
By the way, this is the approach taken in several well-studied applications of tropical geometry (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}).

\subsection{Best Case Analysis and Metric Reasoning}

@@

\subsection{Tropical Polynomials and Log-Probabilities}

>>>>

In this section we illustrate a few directions in which the tropical semantics just introduced could be used to analyze quantitative properties of higher-order programs.

Since algebraic and geometric properties in tropical mathematics are usually more tractable from a computational point of view, in several well-known applications (e.g.~for optimization problems related to machine learning \cite{Pachter2004, Zhang2018, Maragos2021}) one starts from a given model, typically expressed by some polynomial function $f$, and studies 
what properties of the model can be deduced from the \emph{tropicalization} of $f$, noted $\trop f$, i.e.~the transformation of $f$ into a tropical polynomial.

%
%
%several well-known applications of tropical mathematics is to study how much can be deduced of some function starting from the properties of its tropicalization.
%In Section \ref{section5} we will follow a similar direction, investigating what quantitative properties of a higher-order programs are revealed by the study of its tropical interpretation.
%

Here we follow a similar patter: we consider a program $M$, which can be expressed in the form of a polynomial or a power series $f$, and we 
investigate what quantitative properties of $M$ can be deduced from the properties of $\trop f$, that will indeed coincide with the interpretation of $M$ in $\LREL_{!}$.

\input{tropicalization}

\subsection{Best case analysis and metric reasoning}

@@

\subsection{Convergence log-probabilities}
